---
title: "Evolution of gas prices in Llinars del Vallès and Cardedeu"
format: html
---

```{r settings}
#| echo: FALSE

knitr::opts_chunk$set(warning = F, echo = T, message = F)
```

Use this `R` libraries:

```{r libs}
library(httr2)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
```

Define function to retrieve information:

```{r func}
get_gas <- function(filtro, args) {
  paste0(
    "https://sedeaplicaciones.minetur.gob.es",
    "/ServiciosRESTCarburantes/PreciosCarburantes/EstacionesTerrestresHist/",
    filtro, paste(args, collapse = "/")
  ) %>%
    request() %>%
    req_perform() %>%
    resp_body_json() %>%
    .[["ListaEESSPrecio"]] %>%
    map(\(x) as_tibble(list_flatten(x))) %>%
    bind_rows() %>%
    janitor::clean_names()
}
```

Retrieve yesterday's gas prices in catalan provinces:

```{r cat_gas}
prov <- c("08", "43", "17", "25")

gCat <- map(prov, \(x) get_gas(
  filtro = "FiltroProvincia/",
  args = c(format(Sys.Date() - 1, "%d-%m-%Y"), x)
)) %>%
  bind_rows() %>%
  mutate(precio = as.numeric(gsub(",", ".", precio_gasolina_95_e5)))
```

Summary of gas stations and price statistics per province:^[Considering 95 gas - E5]

```{r n_prov}
gCat %>%
  summarise(
    n = n(),
    mean = mean(precio, na.rm = TRUE) %>% round(3),
    median = median(precio, na.rm = TRUE) %>% round(3),
    sd = sd(precio, na.rm = TRUE) %>% round(3),
    .by = provincia
  )
```

We are interest in historic prices from a certain province, for instance Cadiz whose code is 11. Considering daily observations for all Cadiz gas stations, a horizon of one year seems reasonable.

Declare province, product, horizon and API filter parameters:

```{r time-prov}
prov <- "11"
prod <- "01"
h <- 366 / 12
fil <- "FiltroProvinciaProducto/"

cadiz <- map(format(Sys.Date() - 1:h, "%d-%m-%Y"), \(x){
  get_gas(filtro = fil, args = c(x, prov, prod)) %>%
    mutate(fecha = as.Date(x))
}) %>%
  bind_rows() %>%
  mutate(precio = as.numeric(gsub(",", ".", precio_producto)))
```

```{r plot}
gCat %>%
  filter(rotulo %in% (
    summarise(gCat, .by = rotulo, n = n()) %>%
      arrange(n) %>% pull(rotulo) %>% tail(10)
  )) %>%
  drop_na() %>%
  select(rotulo, precio) %>%
  ggplot(aes(rotulo, precio)) +
  geom_boxplot() +
  theme_bw() +
  theme(
    axis.title = element_blank()
  ) +
  labs(
    title = "Gas prices in Catalonia, 2025",
    subtitle = "Euros per litre of gas 95 E5 (10 largest providers)"
  )
```

